---
title: "turnover"
author: "Adam Wawro"
date: "6/11/2020"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include=FALSE}
library(tidyverse)
library(ggridges)
library(viridis)
library(sf)
```

```{r cache=TRUE, include=FALSE}
results_station_clean <- readRDS("../output/clean/results_station_clean.rds")
results_candidate_clean <- readRDS("../output/clean/results_candidate_clean.rds")
results_slate_clean <- readRDS("../output/clean/results_slate_clean.rds")
results_candidate_full <- readRDS("../output/clean/results_candidate_full.rds")
map_powiaty <- readRDS("../output/clean/map_powiaty.rds")
map_wojewodztwa <- readRDS("../output/clean/map_wojewodztwa.rds")
powiaty_table <- readRDS("../output/clean/powiaty_table.rds")
```

## The Sejm elections in Poland in 2019

### Voter turnover

One of the most common ways of displaying regional differences in the voter turnover is displaying them by voivodeship, the first level of administrative subdivisions.

```{r}
turnover_woj <- results_station_clean %>%
  group_by(wojewodztwo) %>%
  summarize(turnover = sum(wydane_karty) / sum(wyborcow_uprawnionych) * 100, .groups = 'drop')

map_wojewodztwa %>%
  left_join(turnover_woj, by = c("nazwa" = "wojewodztwo")) %>%
  ggplot() +
    geom_sf(aes(fill = turnover), size = 0.1, color = "white") +
    scale_fill_viridis(limits = c(46,78), breaks = c(50, 55, 60, 65, 70, 75), name = "turnover / %", option = "E") +
    theme_void() +
    theme(legend.position = "right")
```

This map reveals that the region with the highest turnover is Mazovia [pol. *mazowieckie*], where the capital of Poland is located. The other side of the spectrum is occupied by two peripherally located regions: Opole [pol. *opolskie*] in southern Poland and Warmia-Masuria [pol. *warmińsko-mazurskie*] in north-eastern part of the country.

More interesting patterns are, however, revealed by plotting the turnover over powiats, the subdivisions of voivodeships.

```{r}
turnover_pow <- results_station_clean %>%
  group_by(powiat) %>%
  summarize(turnover = sum(wydane_karty) / sum(wyborcow_uprawnionych) * 100, .groups = 'drop')

map_powiaty %>%
  left_join(turnover_pow, by = c("nazwa" = "powiat")) %>%
  ggplot() +
    geom_sf(aes(fill = turnover)) +
    geom_sf(data = map_wojewodztwa, size = 0.1, fill = NA, color = "white") +
    scale_fill_viridis(limits = c(46,78), breaks = c(50, 55, 60, 65, 70, 75), name = "turnover / %", option = "E") +
    theme_void() +
    theme(legend.position = "right")
```

One can appreciate the bright isles of high turnover, representing major cities and surrounding urban areas. The map suggests that the differences between the voivodeships are not necessarily driven by the ratio of rural to urban areas; clearly the peripheral regions of Mazovia had higher voter turnout than corresponding areas  of its northern neighbor, Warmia-Masuria.

These differences can be shown by plotting the distribution of voter turnovers on the single polling station level. I took the advanatge of the fact that each station is assigned by the electoral commission with 'urban' or 'rural' category. Separating urban and rural areas within each voivodeship gives insight into region-specific differences.

```{r}
results_station_clean %>%
  filter(typ_obszaru %in% c("miasto", "wieś")) %>%
  ggplot(aes(x=wojewodztwo, y=(wydane_karty / wyborcow_uprawnionych * 100), fill = typ_obszaru)) +
  stat_boxplot(outlier.shape = NA) +
  labs(y="turnover / %", x="voivodeship") +
  scale_fill_discrete(name = "area type", labels = c("urban", "rural")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Generally speaking, in all voivodeships urban areas had higher turnover than rural. This graph also shows quantitatively what we suspected in the previous paragraph; the median turnover in rural Mazovia was higher than in urban areas of Opole or Warmia-Masuria.

The same data subset can be looked at from a different perspective; let's see if there is correlation between the turnover and the fraction of voters in urban areas on the powiat level.  

```{r}
turnover_byurbanvoters <- results_station_clean %>%
  filter(typ_obszaru %in% c("miasto", "wieś")) %>%
  group_by(powiat, typ_obszaru) %>%
  summarize(n_voters = sum(wyborcow_uprawnionych), .groups = 'drop') %>%
  pivot_wider(names_from = typ_obszaru, values_from = n_voters) %>%
  replace_na(list(miasto = 0, wieś = 0)) %>%
  mutate(perc_urban = miasto / (miasto + wieś) * 100) %>%
  left_join(turnover_pow, by = "powiat")

ggplot(turnover_byurbanvoters, aes(x = perc_urban, y = turnover)) +
  geom_point() +
  geom_smooth(formula = (y ~ x), method = 'lm') +
  geom_smooth(data = turnover_byurbanvoters[turnover_byurbanvoters$perc_urban<100, ], formula = (y ~ x), method = 'lm', color = "red") +
  labs(y = "turnover / %", x="voters in urban areas / %") +
  expand_limits(y = c(40,80))
```

There is a clear correlation if all powiats are considered (blue line), but it seemed that it might be largely driven by the group of '100% urban voters' powiats, which are in fact 'cities with powiat rights' &ndash; a special category of powiats. Excluding these points completely changes the result (red line): the correlation is no longer observed.

One factor that might affect the willingness to vote is distance to the closest polling station. Although I could not find a summary of areas served by each of the polling stations, a reasonable approximation could be achieved by calculating the average area served by a single polling station for each powiat: powiat area data was pulled from the Polish Wikipedia page.

```{r}
results_station_clean %>%
  filter(typ_obszaru %in% c("miasto", "wieś")) %>%
  group_by(powiat) %>%
  summarize(turnover = sum(wydane_karty) / sum(wyborcow_uprawnionych) * 100, n_stations = n(), n_voters = sum(wyborcow_uprawnionych), .groups = 'drop') %>%
  left_join(powiaty_table, by = "powiat") %>%
  mutate(area_perstation = powierzchnia / n_stations) %>%
  ggplot(aes(x = area_perstation, y = turnover)) +
    geom_point() +
    geom_smooth(formula = (y ~ x), method = 'lm') +
    labs(x = expression("average area per station / km"^2), y = "turnover / %") +
    expand_limits(y = c(40,80))
```

This correlation is clear: the larger average area served by per polling station on a powiat level, the lower the expected turnover.


Another interesting thing to look at is whether the size of a polling station correlates with the voter turnover. We can suspect that the polling stations serving more densely populated areas might be larger, so let's separate urban and rural areas, as well as voivodeships.

```{r, fig.height = 7}
results_station_clean %>%
  filter(typ_obszaru %in% c("miasto", "wieś")) %>%
  ggplot(aes(x = wyborcow_uprawnionych, y = (wydane_karty / wyborcow_uprawnionych * 100))) +
  stat_density2d(data = filter(results_station_clean, typ_obszaru == "miasto"), color = "blue", alpha = 0.5) +
  stat_density2d(data = results_station_clean[results_station_clean$typ_obszaru == "wieś",], color = "red", alpha= 0.5) +
  geom_smooth(data = filter(results_station_clean, typ_obszaru == "miasto"), method = "lm", se = FALSE, color = "blue", size = 0.5) +
  geom_smooth(data = filter(results_station_clean, typ_obszaru == "wieś"), method = "lm", se = FALSE, color = "red", size = 0.5) +
  labs(x = "registered voters per station", y = "turnover / %") +
  expand_limits(y = c(0,100)) +
  facet_wrap("wojewodztwo")
```

Well, our suspicion was correct; urban areas (blue) are indeed on average served by larger polling stations. The size of a polling station is generally positively correlated the expected turnover, with the notable exception of Opole [pol. *opolskie*] voivodeship. It can be also seen that the correlation is consistently stronger in urban areas across all regions.
